---
title: "Notes_datatable"
author: "Alan"
date: "2 de octubre de 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Course Description
The R data.table package is rapidly making its name as the number one choice for handling large datasets in R. This online data.table tutorial will bring you from data.table novice to expert in no time. Once you are introduced to the general form of a data.table query, you will learn the techniques to subset your data.table, how to update by reference and how you can use data.table’s set()-family in your workflow. The course finishes with more complex concepts such as indexing, keys and fast ordered joins. Upon completion of the course, you will be able to use data.table in R for a more efficient manipulation and analysis process. Enjoy!


```{r}
#install.packages("data.table")
library("data.table")

```





## Data.table novice
    Introduction on what exactly a data.table is, how it differs from the traditional data.frame in R, and     understanding the general form of a data.table    query.


### Create and subset a data.table

Welcome to the interactive exercises for your data.table course. Here you will learn the ins and outs of working with the data.table package.

While most of the material is covered by Matt and Arun in the videos, you will sometimes need to show some street smarts to get to the right answer. Remember that before using the hint you can always have a look at the official documentation by typing ?data.table in the console.

Let's start with some warm-up exercises based on the topics covered in the video. Recall from the video that you can use L after a numeric to specify that it is an integer. You can also give columns with different lengths when creating a data.table, and R will "recycle" the shorter column to match the length of the longer one by re-using the first items. In the example below, column x is recycled to match the length of column y:

data.table(x = c("A", "B"), y = 1:4)
   x y
1: A 1
2: B 2
3: A 3
4: B 4
You can also review the slides used in the videos by pressing the slides button.

Instructions
Create a data.table my_first_data_table with a column x = c("a", "b", "c", "d", "e") and a column y = c(1, 2, 3, 4, 5). Use the function data.table().
Create a two-column data.table DT that contains the four integers 1, 2, 1 and 2 in the first column a and the letters A, B, C and D in the second column b. Use recycling so that the contents of a will be automatically used twice. Note that LETTERS[1] returns "A", LETTERS[2] returns "B", and so on.
Select the third row of DT and just print the result to the console.
Select the second and third rows without using commas and print the result to the console.

```{r}
# The data.table package is preloaded
library("data.table")
# Create my_first_data_table
my_first_data_table <- data.table(x = c("a", "b", "c", "d", "e"), 
                                  y = c(1, 2, 3, 4, 5)
                                  )
  
# Create a data.table using recycling
DT <- data.table(a = c(1L, 2L), b = LETTERS[1:4])
DT
# Print the third row to the console
DT[3]

# Print the second and third row to the console without using commas
DT[2:3]

```

## Getting to know a data.table

You can pass a data.table to base R functions like head() and tail() that accept a data.frame because data.tables are also data.frames. Also, keep in mind that the special symbol .N, when used inside square brackets, contains the number of rows. For example, DT[.N] and DT[nrow(DT)] will both return the last row in DT.

Instructions
Select the second to last row of the table using .N.
Return the column names() of the data.table.
Return the number of rows and number of columns of the data.table using the dim() function.
Select row 2 twice and row 3 once, returning a data.table with three rows (two of which are identical).


```{r}
# DT and the data.table package are pre-loaded

# Print the second to last row of DT using .N
DT[.N-1]

# Print the column names of DT
names(DT)

# Print the number or rows and columns of DT
dim(DT)

# Print a new data.table containing rows 2, 2, and 3 of DT
DT[c(2,2,3)]

```



## A data.table of a vector?

A data.table DT is preloaded in your workspace on the right. Type DT in the console to have a look at it. As you have learned in the video, you can select a column from that data.table with DT[, .(B)].

What do you think is the output of DT[, B]?

## A non-existing column

Have a close look at 1.1 and 1.2 from the data.table package FAQs.

Type D <- 5 in the console. What do you think is the output of DT[, .(D)] and DT[, D]?

Well done! Column D does not exist in DT and is thus not seen as a variable. This causes data.table to look for D in DT's parent frame. Also note that .() in j always returns a data.table


## Subsetting data.tables

As a reminder, DT[i, j, by] is pronounced

Take DT, subset rows using i, then calculate j grouped by by.

In the video, the second argument j was covered. j can be used to select columns by wrapping the column names in .().

In addition to selecting columns, you can also call functions on them as if the columns were variables. For example, if you had a data.table heights storing people's heights in inches, you could compute their heights in feet as follows:

    name  eye_color   height_inch
1:   Tom      Brown            69
2: Boris       Blue            71
3:   Jim       Blue            68

heights[, .(name, 
              height_ft = height_inch / 12)]
    name   height_ft
1:   Tom    5.750000
2: Boris    5.916667
3:   Jim    5.666667
Instructions
Create a subset containing the columns B and C for rows 1 and 3 of DT. Simply print out this subset to the console.
From DT, create a data.table, ans with two columns: B and val, where val is the product of A and C.
Fill in the blanks in the assignment of ans2, such that it equals the data.table specified in target. Use columns from the previously defined data.tables to produce the val column.

```{r}
# DT and the data.table package are pre-loaded

DT <- data.table(A = 1:5, B = c("a", "b", "c", "d", "e"), C = 6:10)

# Subset rows 1 and 3, and columns B and C
DT[c(1,3),.(B,C)]

# Assign to ans the correct value
ans <- DT[, .(B, val = A*C)]
ans  
# Fill in the blanks such that ans2 equals target
target <- data.table(B = c("a", "b", "c", "d", "e", 
                           "a", "b", "c", "d", "e"), 
                     val = as.integer(c(6:10, 1:5)))
ans2 <- DT[, .(B, val = c(6:10, 1:5))]
ans2

```


## Data.table yeoman
    Learn how to do multiple operations on the same data.table in one single statement, how to easily take     a subset of your data, update by reference, and work with the data.table set()-family.


### Chaining, the basics

Now that you are comfortable with data.table's DT[i, j, by] syntax, it is time to practice some other very useful concepts in data.table. Here, we'll have a more detailed look at chaining.

Chaining allows the concatenation of multiple operations in a single expression. It's easy to read because the operations are carried out from left to right. Furthermore, it helps to avoid the creation of unnecessary temporary variables (which could quickly clutter one's workspace).

Instructions
In the previous section, you calculated DT2 by taking the cumulative sum of C while grouping by A and B. Next, you selected the last two values of C from DT2 while grouping by A alone. This code is included in the sample code. Use chaining to restructure the code. Simply print out the result of chaining.    
    
```{r}
# The data.table package has already been loaded

# Build DT
DT <- data.table(A = rep(letters[2:1], each = 4L), 
                 B = rep(1:4, each = 2L), 
                 C = sample(8)) 
DT

# Combine the two steps in a one-liner
DT2 <- DT[, .(C = cumsum(C)), by = .(A, B)]
DT2
DT2[, .(C = tail(C, 2)), by = A]
```
    
### Chaining your iris dataset

In the previous chapter, you converted the iris dataset to a data.table DT. This DT is already available in your workspace. Print DT to the console to remind yourself of its contents. Now, let's see how you can use chaining to simplify manipulations and calculations.

Instructions
Get the median of each of the four columns Sepal.Length, Sepal.Width, Petal.Length and Petal.Width, while grouping by Species. Reuse the same column names (e.g. the column containing the median Sepal.Length is still called Sepal.Length). Next, order() Species in descending order using chaining. This is deliberately repetitive, but we have a solution for you in the next exercise!

```{r}
# The data.table DT is loaded in your workspace
DT <- as.data.table(iris)


# Perform chained operations on DT
DT[, .(Sepal.Length = median(Sepal.Length), 
       Sepal.Width = median(Sepal.Width), 
       Petal.Length = median(Petal.Length),
       Petal.Width = median(Petal.Width)), 
   by = Species][order(-Species)]

```

### Programming time vs readability

It is a good idea to make use of familiar functions from base R to reduce programming time without losing readability.

The data.table package provides a special built-in variable .SD. It refers to the subset of data for each unique value of the by argument. That is, the number of observations in the output will be equal to the number of unique values in by.

Recall that the by argument allows us to separate a data.table into groups. We can now use the .SD variable to reference each group and apply functions separately. For example, suppose we had a data.table storing information about dogs:

Sex	Weight	Age	Height
M	40	1	12
F	30	4	7
F	80	12	9
M	90	3	14
M	40	6	12
We could then use

dogs[, lapply(.SD, mean), by = Sex]
to produce average weights, ages, and heights for male and female dogs separately:

   Sex   Weight      Age   Height
1:   M 56.66667 3.333333 12.66667
2:   F 55.00000 8.000000  8.00000
A data.table DT has been created for you and is available in the workspace. Type DT in the console to print it out and inspect it.

Instructions
Get the mean of columns y and z grouped by x by using .SD.
Get the median of columns y and z grouped by x by using .SD.

```{r}
DT <- data.table(x = c(2, 1, 2, 1, 2, 2, 1), y = c(1, 3, 5, 7, 9, 11, 13),                 z = c(2, 4, 6, 8, 10, 12, 14))

# A new data.table DT is available

# Mean of columns
DT[, lapply(.SD, mean) , by= x]

# Median of columns
DT[, lapply(.SD, median) , by= x]

```

### Introducing .SDcols

.SDcols specifies the columns of DT that are included in .SD. Using .SDcols comes in handy if you have too many columns and you want to perform a particular operation on a subset of the columns (apart from the grouping variable columns).

Using .SDcols allows you to apply a function to all rows of a data.table, but only to some of the columns. For example, consider the dog example from the last exercise. If you wanted to compute the average weight and age (the second and third columns) for all dogs, you could assign .SDcols accordingly:

dogs[, lapply(.SD, mean), .SDcols = 2:3]
   Weight Age
1:     56 5.2
While learning the data.table package, you may want to occasionally refer to the documentation. Have a look at ?data.table for more info on .SDcols.

Yet another data.table, DT, has been prepared for you in your workspace. Start by printing it to the console.

Instructions
Calculate the sum of the columns that start with Q, using .SD and .SDcols. Set .SDcols equal to 2:4.
Set .SDcols to be the result of a function call. This time, calculate the sum of columns H1 and H2 using paste0() to specify the .SDcols argument.
Finally, select all but the first row of the groups names 6 and 8, returning only the grp column and the columns that start with Q. Use -1 in i of .SD and use paste0() again. Type desired_result into the console to see what your answer should look like.

```{r}
DT <- data.table(grp=c(6,6,8,8,8), Q1=sample(5,replace=TRUE),
                 Q2=sample(5,replace=TRUE), Q3=sample(5,replace=TRUE),
                 H1=sample(5,replace=TRUE), H2=sample(5,replace=TRUE))
# A new data.table DT is available
DT
# Calculate the sum of the Q columns
DT[,lapply(.SD, sum), .SDcols = 2:4]

# Calculate the sum of columns H1 and H2 
DT[,lapply(.SD, sum), .SDcols = 5:6]

# Select all but the first row of groups 1 and 2, returning only the grp column and the Q columns
DT[, .SD[-1], by = grp, .SDcols = paste0("Q",1:3)]

```

### Mixing it together: lapply, .SD, .SDcols and .N

This exercise is a challenging one, so don't give up! It's important to remember that whenever the j argument is a list (e.g. if it contains .SD or a call to lapply()), a data.table is returned. For example:

dogs[, lapply(.SD, mean), by = sex, .SDcols = c("weight", "age")]
will return a data.table containing average weights and ages for dogs of each sex.

It's also helpful to know that combining a list with a vector results in a new longer list. Lastly, note that when you select .N on its own, it is renamed N in the output for convenience when chaining.

For this exercise, DT, which contains variables x, y, and z, is loaded in your workspace. You must combine lapply(), .SD, .SDcols, and .N to get your call to return a specific output. Good luck!

Instructions
Get the sum of all columns x, y and z and the number of rows in each group while grouping by x. Your answer should be identical to this:
     x x  y  z N
  1: 2 8 26 30 4
  2: 1 3 23 26 3
Get the cumulative sum of column x and y while grouping by x and z > 8 such that the answer looks like this:
     by1   by2 x  y
  1:   2 FALSE 2  1
  2:   2 FALSE 4  6
  3:   1 FALSE 1  3
  4:   1 FALSE 2 10
  5:   2  TRUE 2  9
  6:   2  TRUE 4 20
  7:   1  TRUE 1 13
```{r}
# DT is pre-loaded
DT <- data.table(x=c(2,1,2,1,2,2,1), y=c(1,3,5,7,9,11,13), z=c(2,4,6,8,10,12,14))

# Sum of all columns and the number of rows
DT[,c(lapply(.SD, sum),.N), by = x, .SDcols= c("x","y","z")]

# Cumulative sum of column x and y while grouping by x and z > 8
DT[,lapply(.SD, cumsum), by = .(by1= x, by2 = z>8), .SDcols= c("x","y")]
```
Adding, updating and removing columns

As you now know, := is defined for use in j only, and is used to update data.tables by reference. One way of using := is the LHS := RHS form, where LHS is a character vector of columns (referenced by name or number) you wish to update and RHS is the corresponding value for each column (Note: LHS stands for "left hand side" and RHS stands for "right hand side" in what follows).

For example, the following line multiplies every row of column C by 10 and stores the result in C:

DT[, C := C * 10]
This first exercise will thoroughly test your understanding of := used in the LHS := RHS form. It's time for you to show off your knowledge! A data.table DT has been defined for you in the sample code.

Instructions
Add a column to DT by reference, named Total, that contains sum(B) for each group in column A.
Add 1L to the values in column B, but only in the rows 2 and 4.
Add a new column Total2 that contains sum(B) grouped by A but just over rows 2, 3 and 4.
Remove the Total column from DT.
Use [[ to select the third column as a vector. Simply print it out to the console.
```{r}
# The data.table DT
DT <- data.table(A = letters[c(1, 1, 1, 2, 2)], B = 1:5)
DT
# Add column by reference: Total
DT[, Total:= sum(B), by = A ]
DT
# Add 1 to column B
DT[c(2,4), B:= B + 1L]
DT
# Add a new column Total2
DT[2:4, Total2 := sum(B), by = A]
DT
# Remove the Total column
DT[, Total := NULL]
DT

# Select the third column using `[[`
DT[[3]]
```
### The functional form

You've had practice with using := in the LHS := RHS form. The second way to use := is with functional form:

DT[, `:=`(colA = colB + colC)]
Notice that the := is surrounded by two tick marks! Otherwise data.table will throw a syntax error. It is also important to note that in the generic functional form above, my_fun() can refer to any function, including the basic arithmetic functions. The nice thing about the functional form is that you can get both the RHS alongside the LHS so that it's easier to read.

Time for some experimentation. A data.table DT has been prepared for you in the sample code.

Instructions
Update B with B + 1, add a new column C with A + B, and add a new column D of just 2's.
A variable my_cols has already been defined. Use it to delete these columns from DT.
Finally, delete column D using the column number (2), not its name (D).

```{r}
# A data.table DT has been created for you
DT <- data.table(A = c(1, 1, 1, 2, 2), B = 1:5)

# Update B, add C and D
DT[, `:=`(B = B + 1, C = A + B,D = rep(2))]

# Delete my_cols
my_cols <- c("B", "C")
DT[, (my_cols) := NULL]

# Delete column 2 by number
DT[, c(2):= NULL]
```

### Ready, set(), go!

The set() function is used to repeatedly update a data.table by reference. You can think of the set() function as a loopable, low overhead version of the := operator, except that set() cannot be used for grouping operations. The structure of the set() function looks like this:

set(DT, index, column, value)
The function takes four arguments:

A data.table with the columns you wish to update
The index used in a loop (e.g. the i in for(i in 1:5))
The column or columns you wish to update in the loop
How the column or columns should be updated
In the next two exercises, you will focus on using set() and its siblings setnames() and setcolorder(). You are two exercises away from becoming a data.table yeoman!

Instructions
A data.table DT has been created for you in the workspace. Check it out!
Loop through columns 2, 3, and 4, and for each one, select 3 rows at random and set the value of that column to NA.
Change the column names to lower case using the tolower() function. When setnames() is passed a single input vector, that vector needs to contain all the new names.
Print the resulting DT to the console to see what changed.


```{r}

set.seed(1)
DT <- data.table(A = sample(5,10,TRUE), B = sample(5,10,TRUE),
                 C = sample(5,10,TRUE), D = sample(5,10,TRUE))
# Set the seed
set.seed(1)

# Check the DT that is made available to you
DT

# For loop with set
for(i in 2:4) set(DT,sample(10,3),i, NA)

# Change the column names to lowercase
setnames(DT, tolower(names(DT)))

# Print the resulting DT to the console
DT
```


### The set() family

A summary of the set() family:

set() is a loopable, low overhead version of :=.
You can use setnames() to set or change column names.
setcolorder() lets you reorder the columns of a data.table.
A data.table DT has been defined for you in the sample code.

Instructions
First, add a suffix "_2" to all column names of DT. Use paste0() here.
Next, use setnames() to change a_2 to A2.
Lastly, reverse the order of the columns with setcolorder().

```{r}
# Define DT
DT <- data.table(a = letters[c(1, 1, 1, 2, 2)], b = 1)
DT
# Add a suffix "_2" to all column names
setnames(DT, paste0(names(DT),"_2"))
DT
# Change column name "a_2" to "A2"
setnames(DT, "a_2", "A2")
DT
# Reverse the order of the columns
setcolorder(DT, c("b_2", "A2"))
DT

```

  
## Data.table expert
    Discover the potential behind indexing, followed by generating and using keys. The final part focuses     on fast ordered joins.

### Selecting rows the data.table way

In the video, Matt showed you how to use column names in i to select certain rows. Since practice makes perfect, and since you will find yourself selecting rows over and over again, it'll be good to do a small exercise on this with the familiar iris dataset.

Instructions
Convert the iris dataset to a data.table and store the result as iris.
Select all the rows where Species is "virginica".
Select all the rows where Species is either "virginica" or "versicolor".

```{r}
# The data.table package is pre-loaded

# Convert iris to a data.table
iris <- data.table(iris)

# Species is "virginica"
iris[Species == "virginica"]

# Species is either "virginica" or "versicolor"
iris[Species %in% c("virginica", "versicolor")]
```

### Removing columns and adapting your column names

In the previous exercise, you selected certain rows from the iris data.table based on the column names. Now you have to take your understanding of the data.table package to the next level by using standard R functions and regular expressions to remove columns and change column names. To practice this, you'll do a little manipulation to prepare for the next exercise.

Since regular expressions can be tricky, here is a quick refresher:

Metacharacters allow you to match certain types of characters. For example, . means any single character, ^ means "begins with", and $ means "ends with".
If you want to use any of the metacharacters as actual text, you need to use the \\ escape sequence.
Instructions
Simplify the names of the columns in iris that contain "Sepal." by removing the "Sepal." prefix. Use gsub() along with the appropriate regular expression inside a call to setnames().
Remove the two columns that start with "Petal" from the iris data.table.

```{r}
# iris as a data.table
iris <- as.data.table(iris)

# Remove the "Sepal." prefix
setnames(iris, names(iris), gsub("^Sepal.", "", names(iris)))
iris
# Remove the two columns starting with "Petal"
iris[, grep("^Petal", names(iris)):= NULL]
iris

```

### Understanding automatic indexing

You've been introduced to the rule that "if i is a single variable name, it is evaluated in the calling scope, otherwise inside DT's scope". This is a very important rule if you want to conceptually understand what is going on when using column names in i. Only single columns on the left side of operators benefit from automatic indexing.

The iris data.table with the variable names you updated in the previous exercise is available in your workspace.

Instructions
Select the rows where the area is greater than 20 square centimeters.
Add a new boolean column containing Width * Length > 25 and call it is_large. Remember that := can be used to create new columns.
Select the rows for which the value of is_large is TRUE.

```{r}
# Cleaned up iris data.table
iris

# Area is greater than 20 square centimeters
iris[ Width * Length> 20 ]

# Add new boolean column
iris[,is_large := Width * Length > 25]

# Now large observations with is_large
iris[is_large == TRUE]

```

### Selecting groups or parts of groups

The previous exercise illustrated how you can manually set a key via setkey(DT, A, B). setkey() sorts the data by the columns that you specify and changes the table by reference. Having set a key will allow you to use it, for example, as a super-charged row name when doing selections. Arguments like mult and nomatch then help you to select only particular parts of groups.

Furthermore, two of the instructions will require you to make use of by = .EACHI. This allows you to run j for each group in which each item in i joins too. The last instruction will require you to produce a side effect inside the j argument in addition to selecting rows.

Instructions
A data.table DT has already been created for you with the keys set to A and B.

Select the "b" group without using ==.
Select the "b" and "c" groups, again without using ==.
Select the first row of the "b" and "c" groups using mult.
Use by = .EACHI and .SD to select the first and last row of the "b" and "c" groups.
Extend the previous command to print out the group before returning the first and last row from it. You can use curly brackets to include two separate instructions inside the j argument.

```{r}
# The 'keyed' data.table DT
DT <- data.table(A = letters[c(2, 1, 2, 3, 1, 2, 3)], 
                 B = c(5, 4, 1, 9, 8, 8, 6), 
                 C = 6:12)
setkey(DT, A, B)

# Select the "b" group
DT["b"]

# "b" and "c" groups
DT[c("b","c")]


# The first row of the "b" and "c" groups
DT[c("b","c"), mult="first"]


# First and last row of the "b" and "c" groups
DT[c("b","c"), .SD[c(1,.N)], by= .EACHI]


# Copy and extend code for instruction 4: add printout
DT[c("b","c"), {print(.SD); .SD[c(1,.N)]}, by= .EACHI]
```

```{r}

```

```{r}

```

